---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Load required packages
```{r, results='hide'}
library(knitr)
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data
Load the data and store as data:
```{r}
data <- read.csv("activity.csv")
```

Preprocess by removing NA values:
```{r}
data.clean <- data[with (data, { !(is.na(steps)) } ), ]
```

## What is mean total number of steps taken per day?
Group rows of data and sum by date:
```{r}
by.date <- group_by(data.clean, date)
steps.by.date <- summarise(by.date, total=sum(steps))
steps.by.date
```

Create histogram of the total number of steps taken each day:
```{r}
hist(steps.by.date$total, main="Histogram of Total nmber of steps per day", xlab="Total number of steps per day")
```

```{r}
summary(steps.by.date)
```

**Mean total number of steps per day is 10766 with the median being 10765.**

## What is the average daily activity pattern?

Preprocess data for the plotting of graph
```{r}
steps.by.interval <- aggregate(steps~interval, data, mean)
```

Create a time series plot of the 5-minute interval with average number of steps taken across all days
```{r}
plot(steps.by.interval$interval, steps.by.interval$steps, type='l', 
     main="Mean number of steps across days", xlab="Interval", 
     ylab="Mean number of steps")
```

```{r}
# find row with max of steps
max.steps.row <- which.max(steps.by.interval$steps)

# find interval with this max
steps.by.interval[max.steps.row, ]
```

**The interval 835 has the maximum average value of steps (206.1698).**

## Imputing missing values

Find out how many rows of NAs
```{r}
sum(is.na(data))
```

**The total number of rows with NAs is 2304.**

Replacing the NAs with the mean of 5 minute intervals.
```{r}
data.new<- data
for(i in 1:nrow(data.new)){
  if(is.na(data.new$steps[i])){
    interval.value <- data.new$interval[i]
    steps.value <- steps.by.interval[steps.by.interval$interval == interval.value,]
    data.new$steps[i] <- steps.value$steps
  }
}
```

Group rows of data and sum by date (NEW w/o NAs):
```{r}
steps.by.date.new <- aggregate(steps ~ date, data.new, sum)
head(steps.by.date.new)
```

Create a histogram of total number of steps per day
```{r}
hist(steps.by.date.new$steps, main="Histogram of total number of steps per day", xlab="Total number of steps in a day")
```

```{r}
summary(steps.by.date.new$steps)
summary(steps.by.date$total)
```

**The mean values stay the same with a slight difference in the median values.**

## Are there differences in activity patterns between weekdays and weekends?

Preprocess the data to indicate Weekday or Weekend
```{r}
data.new['day'] <- weekdays(as.Date(data.new$date))
data.new$day[data.new$day  %in% c('Saturday','Sunday') ] <- "weekend"
data.new$day[data.new$day != "weekend"] <- "weekday"
```

```{r}
data.new$day <- as.factor(data.new$day) # convert day from character to factor

# Calculate average steps by interval across all days
steps.by.date.new <- aggregate(steps ~ interval + day, data.new, mean)

# Create a line graph 
qplot(interval, 
      steps, 
      data = steps.by.date.new, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
  facet_wrap(~ day, ncol = 1)
```
